Using high-resolution variant frequencies to empower clinical genome interpretation

### 2016-10-03 manuscript/SupplementaryMaterial.Rmd
Cutting words

>>>>

### The statistical model

The maximum tolerated allele count (AC) was computed as the AC occurring at the upper bound of the one-tailed 95% confidence interval (95%CI AC) for the established maximum credible allele frequency, given the observed allele number (AN). Since the population is drawn without replacement, this would strictly be a hypergeometric distribution, but this can be modeled as binomial as the sample is much smaller than the population from which it is drawn. For ease of computation, we approximate this with a Poisson distribution. In R, this is implemented as `max_ac = qpois(quantile_limit,an*af)`, where `max_ac` is the 95%CI AC, `quantile_limit` is 0.95 (for a one-sided 95%CI), `an` is the observed allele number, and `af` is the maximum credible population allele frequency.

### Defining maximum credible AF for recessive diseases

The prevalence of a recessive condition can be related to the AF of causative variants by:  
*Prevalence =* $\sum$ *(AF of causative alleles in each contributing gene)^2^ x penetrance*  
approximating to:   
*Prevalence = (combined frequency of causative alleles in gene)^2^ x (number of similar genes) x penetrance*  
and expanding to:  
*Prevalence = (max individual AF x 1/maximum allelic contribution)^2^ x 1/maximum genetic contribution x penetrance*

where *maximum genetic contribution* represents the proportion of all cases that are attributable to the gene under evaluation, and *maximum allelic contribution* represents the proportion of cases attributable to that gene that are attributable to an individual variant. The maximum frequency of a recessive disease causing variant in the population was therefore defined as:  
*max credible AF* = $\sqrt(prevalence) \times$ *maximum allelic contribution* $\times \sqrt(maximum genetic contribution) \times 1/\sqrt(penetrance)$

### 2016-10-03 manuscript/manuscript.Rmd
Cutting words

Whole exome and whole genome sequencing have been instrumental in identifying causal variants in Mendelian disease patients `r citep("10.1016/j.ajhg.2015.06.009")`. As every individual harbors ~12,000-14,000 predicted protein-altering variants `r citep("10.1038/nature19057")`, distinguishing disease-causing variants from benign bystanders is perhaps the principal challenge in contemporary clinical genetics. A variant's low frequency in, or absence from, reference databases is now recognised as a necessary, but not sufficient, criterion for variant pathogenicity `r citep(c("10.1038/gim.2015.30","10.1038/nature13127"))`. The recent availability of very large reference databases, such as the Exome Aggregation Consortium (ExAC) `r citep("10.1038/nature19057")` dataset, which has characterised the population allele frequencies of 10 million genomic variants through the analysis of exome sequencing data from over 60,000 humans, provides an opportunity to obtain robust frequency estimates even for rare variants, improving the theoretical power for allele frequency filtering in Mendelian variant discovery efforts.

@@

The maximum frequency of a dominant disease-causing variant in the population was defined as:  
*maximum credible population AF = prevalence* $\times$ *maximum allelic contribution* $\times$ *1/penetrance*

Disease prevalence estimates were obtained from the literature and taken as the highest value reported. A variant penetrance of 0.5 was used for all analyses, as penetrance estimates for individual variants are not widely available. This corresponds to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and is the minimum found when researching other variants/disorders.

@@

The maximum tolerated allele count (AC) was computed as the AC occurring at the upper bound of the one-tailed 95% confidence interval (95%CI AC) for the established maximum credible allele frequency, given the observed allele number (AN). Since the population is drawn without replacement, this would strictly be a hypergeometric distribution, but this can be modeled as binomial as the sample is much smaller than the population from which it is drawn. For ease of computation, we approximate this with a Poisson distribution. In R, this is implemented as `max_ac = qpois(quantile_limit,an*af)`, where `max_ac` is the 95%CI AC, `quantile_limit` is 0.95 (for a one-sided 95%CI), `an` is the observed allele number, and `af` is the maximum credible population allele frequency.

### Application to recessive diseases

The prevalence of a recessive condition can be related to the allele frequency of causative variants by:  
*Prevalence =* $\sum$ *(allele frequency of causative alleles in each contributing gene)^2^ x penetrance*  
approximating to:   
*Prevalence = (combined frequency of causative alleles in gene)^2^ x (number of similar genes) x penetrance*  
and expanding to:  
*Prevalence = (max individual allele frequency x 1/maximum allelic contribution)^2^ x 1/maximum genetic contribution x penetrance*

where *maximum genetic contribution* represents the proportion of all cases that are attributable to the gene under evaluation, and *maximum allelic contribution* represents the proportion of cases attributable to that gene that are attributable to an individual variant. The maximum frequency of a recessive disease causing variant in the population was therefore defined as:  
*max credible allele frequency* = $\sqrt(prevalence) \times$ *maximum allelic contribution* $\times \sqrt(maximum genetic contribution) \times 1/\sqrt(penetrance)$

@@

We used adjusted AC and AN, meaning variant calls with GQ≥20 and DP≥10. More explanation can be found in **Supplementary Methods**.

@@

We utilized the July 9, 2015 release of ClinVar, extracting variants from XML and TXT releases into a single tab-delimited file through use of a Python implementation of vt normalize `r citep("10.1093/bioinformatics/btv112")`, as described previously `r citep("10.1038/nature19057")`. Only variants annotated as pathogenic and non-conflicted were investigated. ExAC counts were determined by matching on chromosome, position, reference, and alternate alleles. For all variants above the proposed maximum tolerated allele count for HCM, all HGMD annotated literature was reviewed and the level of evidence supporting disease pathogenicity was curated according to ACMG criteria `r citep("10.1038/gim.2015.30")`.

@@

We used a cohort of 322 patients recruited to the Royal Brompton Hospital cardiac Biomedical Research Unit with diagnosis of HCM confirmed by cardiac MRI. These samples were sequenced using the IlluminaTruSight Cardio Sequencing Kit `r citep("10.1007/s12265-016-9673-5")` on theIlluminaMiSeq and NextSeq platforms. This study was subject to ethical approval (REC: 09/H0504/104+5) and informed consent was obtained for all subjects. The number of rare variants in _MYBPC3_, _MYH7_ and the six other sarcomeric genes associated with HCM (_TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_) were calculated for this HCM cohort, and for reference population samples from ExAC. Case/control variant frequencies were calculated for all protein altering variants (frameshift, nonsense, splice donor/acceptor, missense and in-frame insertions/deletions), with frequencies and case/control odds ratios calculated separately for non-overlapping ExAC allele frequency bins with the following breakpoints: 1x10^-5^, 5x10^-5^, 1x10^-4^, 5x10^-4^ and 1x10^-3^. Odds Ratios were calculated as OR = (cases with variant / cases without variant) / (ExAC samples with variant / ExAC samples without variant) along with 95% confidence intervals. In the absence of sample-level genotype data for ExAC, the number of samples with a variant was approximated by the total number of variant alleles - i.e. assuming that each rare variant was found in a distinct sample.

@@

All data required to reproduce these analyses is available at <https://github.com/ImperialCardioGenetics/frequencyFilter>. The manuscript was compiled in R, and source code for the analysis, figures and manuscript, are available at the same location. Curated variant interpretations are deposited in ClinVar *Accession & DOI to be added*. ExAC annotations are available at <exac.broadinstitute.org>. Our allele frequency calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, and the source code available at <http://github.com/jamesware/alleleFrequencyApp>.

@@

For each ExAC variant, we defined a "filtering allele frequency" that represents the threshold disease-specific "maximum credible allele frequency" at or below which the disease could not plausibly be caused by that variant. A variant with a filtering allele frequency ≥ the user-defined maximum credible allele frequency for the disease under consideration should be filtered, while a variant with a filtering allele frequency below the maximum credible remains a candidate (Figure 3). It is important to note that the filtering allele frequency is not desease speific and hence can be applied to any disease of interest. This value has been pre-computed for all variants in ExAC (see **Methods** and **Supplementary Methods**), and is available via the ExAC VCF and browser (http://exac.broadinstitute.org).

To assess the efficiency of our approach, we calculated the filtering allele frequency based on 60,206 exomes from ExAC and applied these filters to a simulated dominant Mendelian variant discovery analysis on the remaining 500 exomes (see **Methods**). Filtering at allele frequencies lower than 0.1% can substantially reduce the number of predicted protein-altering variants in consideration, with the mean number of variants per exome falling from 176 at a cutoff of 0.1% to 63 at a cutoff of 0.0001% (Figure 2a). Additionally, we compared the prevalence of variants in HCM genes in cases and controls across the allele frequency spectrum, and computed disease odds ratios for different frequency bins. The odds ratio for disease-association increases markedly at very low allele frequencies (Figure 2b) demonstrating that increasing the stringency of a frequency filter improves the information content of a genetic result.

>>>>

Whole exome and whole genome sequencing have been instrumental in identifying causal variants in Mendelian disease patients `r citep("10.1016/j.ajhg.2015.06.009")`. As every individual harbors ~12,000-14,000 predicted protein-altering variants `r citep("10.1038/nature19057")`, distinguishing disease-causing variants from benign bystanders is perhaps the principal challenge in contemporary clinical genetics. A variant's low frequency in, or absence from, reference databases is recognised as a necessary, but not sufficient, criterion for variant pathogenicity `r citep(c("10.1038/gim.2015.30","10.1038/nature13127"))`. The recent availability of very large reference databases, such as the Exome Aggregation Consortium (ExAC) `r citep("10.1038/nature19057")` dataset, which has characterised the population allele frequencies of 10 million genomic variants through the analysis of exome sequencing data from over 60,000 humans, provides an opportunity to obtain robust frequency estimates even for rare variants, improving the theoretical power for allele frequency filtering in Mendelian variant discovery efforts.

@@

The maximum alelle frequency (AF) of a dominant disease-causing variant in the population was defined as:  
*maximum credible population AF = prevalence* $\times$ *maximum allelic contribution* $\times$ *1/penetrance*

@@

For recessive conditions, the maximum AF is defined as:
*max credible allele frequency* = $\sqrt(prevalence) \times$ *maximum allelic contribution* $\times \sqrt(maximum genetic contribution) \times 1/\sqrt(penetrance)$

where *maximum genetic contribution* represents the proportion of all cases that are attributable to the gene under evaluation, and *maximum allelic contribution* represents the proportion of cases attributable to that gene that are attributable to an individual variant. Full derivation can be found in **Supplementary Methods**.

Disease prevalence estimates were obtained from the literature and taken as the highest value reported. Variant penetrance of 0.5 was used for all analyses, as penetrance estimates for individual variants are not widely available. This corresponds to the reported penetrance of the HCM variant used to illustrate our approach `r citep("10.1161/CIRCRESAHA.109.216291")` and is the minimum found when researching other variants/disorders.

The maximum tolerated allele count (AC) was computed as the AC at the upper bound of the one-tailed 95% confidence interval (95%CI AC) of a Poisson distribution, for the established maximum credible AF, given the observed allele number (AN). In R: `max_ac = qpois(quantile_limit,an*af)`, where `max_ac` is the 95%CI AC, `quantile_limit` is 0.95 (for a one-sided 95%CI), `an` is the observed allele number, and `af` is the maximum credible population AF.

@@

More explanation can be found in **Supplementary Methods**.

We used adjusted AC and AN, meaning variant calls with GQ≥20 and DP≥10.

@@

Pathogenic and non-conflicted variants from XML and TXT releases of ClinVar (July 9, 2015 release) were extracted using a Python implementation of vt normalize `r citep("10.1093/bioinformatics/btv112")`, as described previously `r citep("10.1038/nature19057")`. ExAC counts were determined by matching on chromosome, position, reference, and alternate alleles. For all variants above the proposed maximum tolerated allele count for HCM, HGMD literature was reviewed and the level of evidence supporting pathogenicity was curated according to ACMG criteria `r citep("10.1038/gim.2015.30")`.

@@

322 HCM patients (confirmed by cardiac MRI) recruited to the Royal Brompton Hospital cardiac BRU were sequenced using the IlluminaTruSight Cardio Sequencing Kit `r citep("10.1007/s12265-016-9673-5")` on Illumina MiSeq and NextSeq platforms. This study had ethical approval (REC: 09/H0504/104+5) and informed consent was obtained for all subjects. The number of rare variants in _MYBPC3_, _MYH7_ and the six other sarcomeric genes associated with HCM (_TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_) were calculated for this HCM cohort, and for reference population samples from ExAC. Variant frequencies were calculated for all protein altering variants (frameshift, nonsense, splice donor/acceptor, missense and in-frame insertions/deletions), with case/control odds ratios calculated separately for non-overlapping ExAC AF bins with the following breakpoints: 1x10^-5^, 5x10^-5^, 1x10^-4^, 5x10^-4^ and 1x10^-3^. Odds Ratios were calculated as OR = (cases with variant/cases without variant) / (ExAC samples with variant/ExAC samples without variant). In the absence of sample-level genotype data for ExAC, the number of samples with a variant was approximated by the total number of variant alleles (assuming each rare variant was from a distinct sample).

@@

All data required to reproduce these analyses is available at <https://github.com/ImperialCardioGenetics/frequencyFilter>. The manuscript was compiled in R, and source code for the analysis, figures and manuscript, are available at the same location. Curated variant interpretations are deposited in ClinVar *Accession & DOI to be added*. ExAC annotations are available at <exac.broadinstitute.org>. Our allele frequency calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, with source code available at <http://github.com/jamesware/alleleFrequencyApp>.

@@

For each ExAC variant, we defined a "filtering allele frequency" that represents the threshold disease-specific "maximum credible allele frequency" at or below which the disease could not plausibly be caused by that variant. A variant with a filtering allele frequency≥the user-defined maximum credible allele frequency for the disease under consideration should be filtered, while a variant with a filtering allele frequency below the maximum credible remains a candidate (Figure 3). The filtering allele frequency is not desease speific and hence can be applied to any disease of interest. This value has been pre-computed for all variants in ExAC (see **Methods** and **Supplementary Methods**), and is available via the ExAC VCF and browser (http://exac.broadinstitute.org).

To assess the efficiency of our approach, we calculated the filtering allele frequency on 60,206 exomes from ExAC and applied these filters to a simulated dominant Mendelian variant discovery analysis on the remaining 500 exomes (see **Methods**). Filtering at allele frequencies lower than 0.1% substantially reduces the number of predicted protein-altering variants in consideration, with the mean number of variants per exome falling from 176 to 63 at cutoffs of 0.1% and 0.0001% respectively (Figure 2a). Additionally, we compared the prevalence of variants in HCM genes in cases and controls across the allele frequency spectrum, and computed disease odds ratios for different frequency bins. The odds ratio for disease-association increases markedly at very low allele frequencies (Figure 2b) demonstrating that increasing the stringency of a frequency filter improves the information content of a genetic result.
